---
title: "Plots"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(magrittr)
require(kableExtra)
```

# Intro
This document illustrates the process used to create Figures 2-4 in from the paper *Genetic and morphological estimates of androgen exposure predict social deficits in multiple neurodevelopmental disorder cohorts.* These figures and data relate only to the study's DevGenes cohort, as raw data from the other cohort --- SPARK -- cannot be publicly released. SPARK data may be obtained by researchers approved by the Simons Foundation Autism Research Initiative, through SFARI Base: https://www.sfari.org/resource/sfari-base. 

All code referenced in this document can be found in either `helper_functions.R` or `devGenes_final_figures_code.R`, and the data on which they are based can be found in `table_S1_final.txt` and \newline `devGenes_final_factor_model_loadings.txt`. 

\newpage
# Helper Functions
## `img()`: A wrapper to make image() more convenient
```{r}
img = function(x,ylab,xlab,axes,col,na.zero=F,breaks,do.breaks=T,...){
 cc = colorRampPalette(c("black","chartreuse"))
 if(any(x<0) & any(x>0)){
	cc = colorRampPalette(c("royalblue","royalblue4","black","orangered","goldenrod1"))
 }
 if(missing(ylab)) ylab=""
 if(missing(xlab)) xlab=""
 if(missing(axes)) axes=F
 if(missing(col)) col=cc(256)
 if(na.zero) x[is.na(x)] = 0
 if(do.breaks & missing(breaks)){
  mx = max(abs(x))
  qt = max(abs(quantile(x,c(0.01,0.99))))
  bk = c(-1*mx,seq(-1*qt,qt,length.out=255),mx)
  image(0:ncol(x),0:nrow(x),t(x),ylim=c(nrow(x),0),ylab=ylab,   
   xlab=xlab,axes=axes,col=col,breaks=bk,...)
 }else{
 image(0:ncol(x),0:nrow(x),t(x),ylim=c(nrow(x),0),ylab=ylab,   
  xlab=xlab,axes=axes,col=col,breaks=breaks,...)
 }
 box()
}
```

\newpage
## `bootlM()` and `lboot()`: wrappers for bootstrapping to get confidence intervals
```{r}
bootlM = function(x,y){
 keep = !is.na(x) & !is.na(y)
 x0 = x[keep]
 y0 = y[keep]
 these = sample(length(x0),replace=T)
 x0 = x0[these]
 y0 = y0[these]

 approx(lowess(x0,y0),xout=seq(0,80,1),rule=2)$y
}

### with input x and y, generate a lowess trend with 95% confidence intervals
lboot = function(x,y,subset,nboot=1000,return.boot=F){
 if(length(x)!=length(y)) stop("x and y have different lengths")
 if(missing(subset)){
  subset=1:length(x)
 }
 x = x[subset]
 y = y[subset]

 ### generate nboot lowess fits
 b = replicate(nboot,bootlM(x,y))

 qu = apply(b,1,quantile,0.975)
 ql = apply(b,1,quantile,0.025)
 qm = rowMeans(b)
 xq = 0:80

 out = list()
 out$x = xq
 out$y = qm
 out$ci = list(upr=qu,lwr=ql)
 out$points = list(x=x,y=y)
 if(return.boot){
  out$boot = b
 }
 return(out)
}

```

\newpage
## `plotl()`: A function to plot the output of `lboot()`
```{r}
plotl = function(obj,pt.col,l.col,shade.col,plot.points=F,xlim=c(0,80),
	ylim,add=F,plot.ci=T,...){
 if(missing(ylim)){
  ylim=range(obj$y,na.rm=T)
  ylim = (c(-1,1)*abs(diff(ylim))) + ylim
 }
 if(!add){
  plot(obj,xlim=xlim,ylim=ylim,xlab="age (years)",type='l',col=l.col,...)
  if(plot.ci){
   polygon(c(obj$x,rev(obj$x)),c(obj$ci$lwr,rev(obj$ci$upr)),col=shade.col,border=l.col)
  }
 }else{
  lines(obj,col=l.col,...)
  if(plot.ci){
   polygon(c(obj$x,rev(obj$x)),c(obj$ci$lwr,rev(obj$ci$upr)),col=shade.col,border=l.col)
  }
 }
 if(plot.points){
  points(obj$points,pch=16,col=pt.col)
 }

}
```

\newpage
# The Data
```{r, echo = FALSE}
l = read.table("devGenes_final_factor_model_loadings.txt",sep="\t",header=T,
	stringsAsFactors=F,row.names=1)
kbl(l, booktabs = T, caption = 'Factor Loadings') %>% 
  kable_styling(latex_options = c("striped", "scale_down", "hold_position")) 
```

```{r, echo = FALSE}
dat = read.table("table_S1_final.txt",sep="\t",header=T,stringsAsFactors=F)

print_dat <- dat[,2:35]
print_dat$morePRSes <- "..."
ellipses <- function(df, n=5L) {
    stopifnot("data.frame" %in% class(df))
    els <- rep("...", ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    out <- df %>% 
        head(n) %>% 
        lapply(as.character) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble() %>% 
        dplyr::bind_rows(els)
    return(out)
}
kbl(ellipses(print_dat, n = 25), 
    booktabs = T, caption = 'Supplementary Table S1') %>% 
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  landscape()
```

# Figure 2: facial and digital masculinity by age
```{r, fig.width = 12, fig.height = 8, message = FALSE, warning = FALSE}
### lboot() returns the lowess trend of the 
### data as well as its 95% CI; the object returned by lboot()
### can be visualized using plotl() as shown below

set.seed(1283)
### get the trends for TD males
idx = !dat$affected & dat$sex_male==1
tdM_flm = lboot(dat$ageyears,dat$facial_masc_raw,subset=idx,return.boot=T)
tdM_drm = lboot(dat$ageyears,dat$digit_ratio_raw,subset=idx,return.boot=T)

### get the trends for TD females
idx = !dat$affected & dat$sex_male==0
tdF_flm = lboot(dat$ageyears,dat$facial_masc_raw,subset=idx,return.boot=T)
tdF_drm = lboot(dat$ageyears,dat$digit_ratio_raw,subset=idx,return.boot=T)

set.seed(7751)
### get the trends for affected males
idx = dat$affected & dat$sex_male==1
affM_flm = lboot(dat$ageyears,dat$facial_masc_raw,subset=idx,return.boot=T)
affM_drm = lboot(dat$ageyears,dat$digit_ratio_raw,subset=idx,return.boot=T)

### get the trends for affected females
idx = dat$affected & dat$sex_male==0
affF_flm = lboot(dat$ageyears,dat$facial_masc_raw,subset=idx,return.boot=T)
affF_drm = lboot(dat$ageyears,dat$digit_ratio_raw,subset=idx,return.boot=T)
```

\newpage
## TD male and female trends for digit ratio
```{r, fig.width = 12, fig.height = 8, message = FALSE, warning = FALSE}
plotl(tdM_drm,l.col="blue",shade.col=rgb(0,0,1,0.2),pt.col=rgb(0,0,1,0.2),
plot.points=T,ylab="2D:4D ratio",ylim=c(0.85,1.1))
plotl(tdF_drm,l.col="red",shade.col=rgb(1,0,0,0.2),pt.col=rgb(1,0,0,0.2),
plot.points=T,ylab="2D:4D ratio",ylim=c(0.85,1.1),add=T)
```

\newpage
## TD male and AFF male trends for digit ratio
```{r, fig.width = 12, fig.height = 8, message = FALSE, warning = FALSE}
plotl(tdM_drm,l.col="grey",shade.col=rgb(0.2,0.2,0.2,0.2),pt.col=rgb(0.2,0.2,0.2,0.2),
plot.points=F,ylab="2D:4D ratio",ylim=c(0.85,1.1))
plotl(affM_drm,l.col="blue",shade.col=rgb(0,0,1,0.2),pt.col=rgb(0,0,1,0.2),
plot.points=T,ylab="2D:4D ratio",ylim=c(0.85,1.1),add=T,plot.ci=F)
### look at the empirical p value for Ha: aff > masc. than TD
### (for digit ratio, lower value is more masculine)
sum(colSums(tdM_drm$b)<sum(affM_drm$y))/1000 ### this is the empirical p-value
```

\newpage
## TD female and AFF female trends for digit ratio
```{r, fig.width = 12, fig.height = 8, message = FALSE, warning = FALSE}
plotl(tdF_drm,l.col="grey",shade.col=rgb(0.2,0.2,0.2,0.2),pt.col=rgb(0.2,0.2,0.2,0.2),
plot.points=F,ylab="2D:4D ratio",ylim=c(0.85,1.1))
plotl(affF_drm,l.col="red",shade.col=rgb(1,0,0,0.2),pt.col=rgb(1,0,0,0.2),
plot.points=T,ylab="2D:4D ratio",ylim=c(0.85,1.1),add=T,plot.ci=F)
### look at the empirical p value for Ha: aff > masc. than TD
### (for digit ratio, lower value is more masculine)
sum(colSums(tdF_drm$b)<sum(affF_drm$y))/1000 ### this is the empirical p-value
```

\newpage
## TD male and female trends for facial masculinity
```{r, fig.width = 12, fig.height = 8, message = FALSE, warning = FALSE}
plotl(tdM_flm,l.col="blue",shade.col=rgb(0,0,1,0.2),pt.col=rgb(0,0,1,0.2),
plot.points=T,ylab="facial masculinity",ylim=c(-6,6))
plotl(tdF_flm,l.col="red",shade.col=rgb(1,0,0,0.2),pt.col=rgb(1,0,0,0.2),
plot.points=T,ylab="facial masculinity",ylim=c(-6,6),add=T)
```

\newpage
## TD male and AFF male trends for facial masculinity
```{r, fig.width = 12, fig.height = 8, message = FALSE, warning = FALSE}
plotl(tdM_flm,l.col="grey",shade.col=rgb(0.2,0.2,0.2,0.2),pt.col=rgb(0.2,0.2,0.2,0.2),
plot.points=F,ylab="facial masculinity",ylim=c(-6,6))
plotl(affM_flm,l.col="blue",shade.col=rgb(0,0,1,0.2),pt.col=rgb(0,0,1,0.2),
plot.points=T,ylab="facial masculinity",ylim=c(-6,6),add=T,plot.ci=F)
### look at the empirical p value for Ha: aff > masc. than TD
sum(colSums(tdM_flm$b)>sum(affM_flm$y))/1000 ### this is the empirical p-value
```

\newpage
## TD female and AFF female trends for facial masculinity
```{r, fig.width = 12, fig.height = 8, message = FALSE, warning = FALSE}
plotl(tdF_flm,l.col="grey",shade.col=rgb(0.2,0.2,0.2,0.2),pt.col=rgb(0.2,0.2,0.2,0.2),
plot.points=F,ylab="facial masculinity",ylim=c(-6,6))
plotl(affF_flm,l.col="red",shade.col=rgb(1,0,0,0.2),pt.col=rgb(1,0,0,0.2),
plot.points=T,ylab="facial masculinity",ylim=c(-6,6),add=T,plot.ci=F)
### look at the empirical p value for Ha: aff > masc. than TD
sum(colSums(tdF_flm$b)>sum(affF_flm$y))/1000 ### this is the empirical p-value
```

\newpage


# Figure 3: Relationships with diagnosis and parent-reported problems

## Top: Boxplots showing  masculinity scores distribution and barplots showing significance
```{r}
### pull out the Dx indicators
dxlab = c("adhd","asd","ID","lang","epilepsy","depression","bipolar","anxiety")
dx = as.data.frame(sapply(dat[,dxlab],as.integer))
rownames(dx) = rownames(dat)

cc = c("grey",rep("orangered",5),rep("mistyrose",3))

### structure the masculinity data according to Dx
ll = lapply(dx,function(x) dat$Zdrm[x==1])
ll = c(list(TD=dat$Zdrm[rowSums(dx)==0]),ll)
ll_drm = ll
ll = lapply(dx,function(x) dat$Zflm[x==1])
ll = c(list(TD=dat$Zflm[rowSums(dx)==0]),ll)
ll_flm = ll

## t-tests comparing TD to all other Dxs
td_vs_dx = function(x) {
 p = sapply(x[-1],function(y) t.test(x[[1]],y)$p.v)
 fdr = p.adjust(p,'fdr')
 out = cbind(p,fdr)
 rownames(out) = names(x)[-1]
 return(out)
}

p_drm = td_vs_dx(ll_drm)
p_flm = td_vs_dx(ll_flm)
```

```{r, fig.width = 12, fig.height = 6}
layout(matrix(c(1,3,2,4),2,2),heights=c(1,5))
par(mar=c(0,4,1,2))
  barplot(-log10(t(rbind(c(NA,NA),p_drm))),
  beside=T,
  las=2,
  ylim=c(0,2.5),
  ylab="-log10 P or FDR",
  names.arg=rep("",9),
  legend.text=c("p","FDR"),
  args.legend=list(x="topleft",bty="n",border=NA,inset=0.03))
  abline(h=-log10(0.05),col='red',lty=2)
  barplot(-log10(t(rbind(c(NA,NA),p_flm))),
  beside=T,
  las=2,
  ylim=c(0,2.5),
  ylab="-log10 P or FDR",
  names.arg=rep("",9),
  legend.text=c("p","FDR"),
  args.legend=list(x="topright",bty="n",border=NA,inset=0.03))
  abline(h=-log10(0.05),col='red',lty=2)
 
## the boxplots showing the distribution of the masculinity scores within
par(mar=c(5,4,0.5,2))
  bp = boxplot(ll_drm,las=2,ylim=c(-4,4),
  col=c("grey",rep("white",8)),
  ylab="DRM (age and sex-corrected)",
  xaxs="i")
  ### pattern fills to emphasize (non)significance
  rect((2:9)-.4, 
  bp$stats[2,-1], 
  (2:9)+.4, 
  bp$stats[4,-1],
  density=c(rep(30,8)), ## tweaked according to which comparisons are sig.
  col=cc[-1],border="black") 
  bp = boxplot(ll_flm,las=2,ylim=c(-4,4),
  col=c("grey",rep("orangered",3),rep("white",5)),
  ylab="FLM (age and sex-corrected)",
  xaxs="i")
  ### pattern fills to emphasize (non)significance
  rect((2:9)-.4, 
  bp$stats[2,-1], 
  (2:9)+.4, 
  bp$stats[4,-1],
  density=c(0,0,0,rep(30,6)), ## tweaked according to which comparisons are sig.
  col=cc[-1],border="black") 
```

\newpage
## Bottom: Parent-report Factors
```{r, fig.width = 12, fig.height = 6}
# Xs is the matrix of parent-reported items (we are only distributing factor scores)
# fac = factanal(scale(Xs),factors=11,scores="regression")

### grab the factor scores from the supplemental table
S =  dat[,grep("Factor",colnames(dat))]
### look at the correlations
cors = cor(dat[,c("Zdrm","Zflm")],S,use="pairwise.complete.obs",method="spearman")

 ### set layout
 layout(matrix(c(1,2,3),1,3),widths=c(1.5,5,5))
 ### set colors, labels, and breaks for loadings
 cp = colorRampPalette(c("steelblue","ivory","red"))
 flab = c("F1:ability","F2:sens/order","F3:aggression","F4:feeding",
	"F5:self-harm","F6:stimming","F7:social anxiety","F8:cutting",
	"F9:social/friends","F10:GI","F11:sound sensitivity")
 bk = c(-1,seq(-1,-0.1,length.out=127),0,seq(0.1,1,length.out=127),1)

 ### the key for the factor loadings
 par(mar=c(5,7,12,1))
 img(as.matrix(seq(1,-1,length.out=256)),col=cp(256),breaks=bk)
 axis(2,at=c(0,256/2,256),labels=c(1,0,-1))
 mtext("factor loadings",side=2,at=256/0,line=3,cex=0.8)

 ### the factor loadings
 par(mar=c(5,1,12,1))
 img(t(l),breaks=bk,col=cp(256))
 mtext(rownames(l),side=3,at=c(1:nrow(l))-0.5,las=2,line=0.5,cex=0.5)
 abline(h=c(0:12),col='grey')
 box()

 ### the barplot of correlations w/ factor scores
 par(mar=c(5,9,12,2))
 bp = barplot(cors[2:1,11:1],beside=T,las=2,yaxs="i",names.arg=flab[11:1],axes=F,
	ylab="",col=c("darkseagreen","cadetblue"),
	legend.text=c("FLM","DRM"),
	args.legend=list(x="topleft",inset=0.03,border=NA,bty='n',density=-1),
	border=NA,xlim=c(-0.2,0.22),
	density=rev(c(rep(30,16),-1,-1,rep(30,4))),horiz=T)
 axis(1,line=1.5)
 mtext("Spearman's rho",side=1,at=0.025,line=4,cex=0.8)
 text(c(0.21,0.21),bp[1:2,3],rep("*",2),col='red',cex=2)
```

\newpage
### Bottom: Statistics
```{r, echo = FALSE}
apply(S,2,function(y){cor.test(y,dat$Zdrm,method="spearman")}) %>% 
  lapply(broom::tidy) %>%
  tibble::enframe() %>%
  tidyr::unnest(value) %>%
  dplyr::mutate(FDR = p.adjust(p.value, 'fdr')) %>%
  dplyr::select(Factor = name, '$\\rho$' = estimate, FDR) %>%
  kbl(booktabs = T, 
      format = 'latex', 
      caption = 'Digit Ratio Masculinity', 
      escape = FALSE) %>% 
    kable_styling(latex_options = c("striped", "hold_position")) %>%
    row_spec(9, bold = T)

apply(S,2,function(y){cor.test(y,dat$Zflm,method="spearman")}) %>%
  lapply(broom::tidy) %>%
  tibble::enframe() %>%
  tidyr::unnest(value) %>%
  dplyr::mutate(FDR = p.adjust(p.value, 'fdr')) %>%
  dplyr::select(Factor = name, '$\\rho$' = estimate, FDR) %>%
  kbl(booktabs = T, 
      format = 'latex', 
      caption = 'Facial Landmark Masculinity', 
      escape = FALSE) %>% 
    kable_styling(latex_options = c("striped",  "hold_position")) %>%
    row_spec(9, bold = T)
```

\newpage
# Figure 4: PRS in devGenes

## Left: PRS Associations
```{r}
### pull out the PRS and covariates from the main table for convenience
prs = dat[,grep("p0",colnames(dat))]
covars = dat[,c(2:3,grep("PC",colnames(dat)))]
covars[[1]] = log(covars[[1]])

### wrapper to grab the PRS-related effect (x) and associated statistics
### from a linear model
fsig = function(x,y) summary(lm(y~.,data=data.frame(covars,x)))$coef[9,]

cfsoc = t(apply(prs,2,fsig,S[,9]))
cfdrm = t(apply(prs,2,fsig,dat$Zdrm))
cfflm = t(apply(prs,2,fsig,dat$Zflm))

### add FDR
cfsoc = cbind(cfsoc,FDR = p.adjust(cfsoc[,4],'fdr'))
cfdrm = cbind(cfdrm,FDR = p.adjust(cfdrm[,4],'fdr'))
cfflm = cbind(cfflm,FDR = p.adjust(cfflm[,4],'fdr'))

### assemble the t-statistics and arrange for the barplot
p1 = cbind(drm=cfdrm[grepl("0\\.1$",rownames(cfdrm)),3],
	flm=cfflm[grepl("0\\.1$",rownames(cfflm)),3],
	F9=cfsoc[grepl("0\\.1$",rownames(cfsoc)),3])
p01 = cbind(drm=cfdrm[grepl("0\\.01$",rownames(cfdrm)),3],
	flm=cfflm[grepl("0\\.01$",rownames(cfflm)),3],
	F9=cfsoc[grepl("0\\.01$",rownames(cfsoc)),3])

### this is what will be plotted
pdrm = cbind(p01[,1],p1[,1])
pflm = cbind(p01[,2],p1[,2])
psoc = cbind(p01[,3],p1[,3])

### helper function to pattern fill according to significance
fdens = function(x) ifelse(t(abs(x)>=1.96),-1,30)
### define colors and labels
cc = c("steelblue1","steelblue4")
plab = c("UKBB\ntestosterone","UKBB\nSHBG","UKBB\nsoc. dissatis.","PGC\nASD","PGC\nADHD","UKBB\nCog","SSGAC\nEA")
```

\newpage
### Digit Ratio Masculinity
```{r, fig.height = 6, fig.width = 8}
par(mar=c(1,5,2,2))
bp = barplot(t(pdrm),
  beside=T,
  las=2,
  ylim=c(-4.2,4),
  names.arg=rep("",7),
  density=fdens(pdrm),
  col=cc,border=NA,
  legend.text=c("Psnp<0.01","Psnp<0.1"),
  args.legend=list(x="topright",bty='n',border=NA,density=-1),
  ylab="DRM vs. PRS association\n(t-statistic)")
  rect(t(bp)[,1]-1,-4.20,t(bp)[,2]+1,-3.55)
  text(bp[1,]+0.5,rep(-3.88,7),plab,cex=0.85)
  abline(h=c(-1.96,1.96),col='red',lty=2)
```

\newpage
### Facial Landmark Masculinity
```{r, fig.height = 6, fig.width = 8}
par(mar=c(1,5,2,2))
barplot(t(pflm),
  beside=T,
  las=2,
  ylim=c(-4.2,4),
  names.arg=rep("",7),
  density=fdens(pflm),
  col=cc,border=NA,
  ylab="FLM vs. PRS association\n(t-statistic)")
  abline(h=c(-1.96,1.96),col='red',lty=2)
  rect(t(bp)[,1]-1,-4.20,t(bp)[,2]+1,-3.55)
  text(bp[1,]+0.5,rep(-3.88,7),plab,cex=0.85)
```

\newpage
### Factor 9 (social)
```{r, fig.height = 6, fig.width = 8}
par(mar=c(1,5,2,2))
barplot(t(psoc),
  beside=T,
  las=2,
  ylim=c(-4.2,4),
  names.arg=rep("",7),
  density=fdens(psoc),
  col=cc,border=NA,
  ylab="F9 (social/friends) vs. PRS association\n(t-statistic)")
  abline(h=c(-1.96,1.96),col='red',lty=2)
  rect(t(bp)[,1]-1,-4.20,t(bp)[,2]+1,-3.55)
  text(bp[1,]+0.5,rep(-3.88,7),plab,cex=0.85)
```

\newpage
## Right: Categorical PRS Comparison
```{r, fig.width = 7, fig.height =12}
## set up colors, layout, and labels
par(mfrow=c(3,2),mar=c(4,4,3,2))
cc1 = colorRampPalette(c("ivory","orangered1"))(3)
cc2 = colorRampPalette(c("ivory","paleturquoise3"))(3)
nn = c("bottom 20%","middle 60%","top 20%")

 boxplot(dat$Zdrm~prs[,1],
	names=nn,
	ylab="DRM",
	main="testosterone PRS",
	col=cc1)
 boxplot(dat$Zdrm~prs[,9],
	names=nn,
	ylab="DRM",
	main="SHBG PRS",
	col=cc2)
 boxplot(dat$Zflm~prs[,1],
	names=nn,
	ylab="FLM",
	main="testosterone PRS",
	col=cc1)
 boxplot(dat$Zflm~prs[,9],
	names=nn,
	ylab="FLM",
	main="SHBG PRS",
	col=cc2)
 boxplot(S[,9]~prs[,1],
	names=nn,
	ylab="Factor 9 (social)",
	main="testosterone PRS",
	col=cc1)
 boxplot(S[,9]~prs[,9],
	names=nn,
	ylab="Factor 9 (social)",
	main="SHBG PRS",
	col=cc2)
```


